---
title: Introduction to Analyzing and Visualizing Data with R
output: html_document
---

# About R Markdown

*To be done*.

https://rmarkdown.rstudio.com/

RStudio and R Markdown are a great combination for working, keeping notes, reproducible research and collaborating.  

(You may also want to look at Jupyter Notebooks (https://jupyter.org/), which can be used with R, Julia or Python.)

# Analyzing data

## Setup

``` {r Setup}
## Anything starting with ## is a comment, not code
## Remove them if you want to run the line

## install.packages("tidyverse") ## If not already done
library(tidyverse) ## Will load ggplot2 and others
```

# What is the Tidyverse?

Indeed, what?

## Get the data (surveys)

Use # to comment in scripts and R markdown.

``` {r Load data}
# Load the data for this lesson by using the read_csv function (from the tidyverse) and assigning the dataset to an object named "surveys"
surveys <- read_csv("http://kbroman.org/datacarp/portal_data_joined.csv")
```

## Get the data (GHG)

``` {r}
## Download the data
ghg_raw <- read_csv("https://raw.githubusercontent.com/yorkulibraries/rworkshop/master/ghg-cleaned.csv")

## View the data
## RStudio: click on variable name in upper right
ghg_raw


## We’re using the pipe and the tidyr package here,
## but consider it magic for now.
ghg <- ghg_raw %>% pivot_longer(-year, names_to = "sector", values_to = "emissions")

## View it now
## RStudio: click on variable name in uppper right
ghg
```


## Using R like a calculator

``` {r}
# Using R like a calculator
2020 - 1971
2020 - 1946

8.95 * 1.13
```

## Inspecting data frames

``` {r}
# head: shows the top few rows
head(surveys)
```

``` {r}
# tail: shows the bottom few rows
tail(surveys)
```

``` {r}
# str shows the structure
str(surveys)
```

##  Challenge #1

Study the output of str(surveys). How are missing values being treated?

Other useful ways of inspecting data frames:

dim(surveys)
ncol(surveys)
nrow(surveys)
names(surveys)
colnames(surveys)
rownames(surveys)
summary(surveys)

## Indexing

Data frames have rows and columns, and there are various ways to get at one or the other or a specific cell.  When indexing, rows go first and columns second: you count down and then over.

``` {r}
surveys[1, 1]
surveys[1, 7]
surveys[10001, 7]
surveys[2, ]
surveys[2,]  # don't _need_ the spaces
surveys[,7]  # Helpful things with tibbles and RStudio make it easier to inspect this
```

## Vectors

``` {r}
sex <- surveys[,7]
sex <- surveys[, "sex"]
sex <- surveys$sex
sex <- surveys[["sex"]]
```

``` {r}
# Pull out a single value from a vector
sex[1]
sex[10001]
```

``` {r}
# Create a vector 
c(1, 4, 6)
```

``` {r}
# Pull out multiple values
sex[c(1,4,6)] # Why are 4 and 6 not seen?
```

``` {r}
sex[c(1, 10001)] # <- pull out 1st and 10001th.
```

``` {r}
# More ways to create vectors
1:10
10:1
sex[1:10]
```

#every 2nd value
seq(1, 10, by=2)
sex[seq(1,10, by=2)]

#seq function
?seq # help
seq(1, 11)
seq(1, 11, 2)
seq(1, 11, by=2)
seq(to=11, from=1, by=2)

#first ten rows of surveys
surveys[1:3,]
surveys[10001:10003,]
surveys[5:7, 1:7]

## Challenge 2

The function nrow() on a data.frame returns the number of rows.

Use nrow(), in conjuction with seq() to create a new data frame called surveys_by_10 that includes every 10th row of the survey data frame starting at row 10 (10, 20, 30, .)

``` {r}
# How would you  make sequence that goes up by tens?
# How would you use that to extract just those rows from surveys?
# Try some code here.  Edit and rerun as you work on it.

```

## Dealing with missing data

``` {r}
heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
na.omit (heights)
```

surveys_noblanks <- read.csv("http://kbroman.org/datacarp/portal_data_joined.csv", na.strings="")

###dplyr

#load the dplyr package
library(dplyr)

#select some columns
selected_col <- select(surveys, sex, species_id, plot_type, weight)
head(selected_col)

#filter out some rows
selected_row <- filter(surveys, year == 2002)
head(selected_row)

#filter out some rows
selected_row <- filter(surveys, year == 2002, weight>78)
head(selected_row)

selected_row <- filter(surveys, sex == "F", weight>78)
head(selected_row)

#pipe operator
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

selected_rows <- filter(surveys, weight<5)
result <- select(selected_rows, species_id, sex, weight)

# Challenge 3

Using pipes, subset the data to include individuals collected before 1995, and retain the columns year, sex, and weight.

# Visualizing data

## Introduction

The goal here is to show how the [ggplot2](https://ggplot2.tidyverse.org/) system works (combining commands to build up a chart piece by piece), to show a variety of examples, and to recreate an existing chart in R.

## Data: Canadian greenhouse gas emissions

Here we will use [greenhouse gas emissions data from Environment and Climate Change Canada](https://www.canada.ca/en/environment-climate-change/services/environmental-indicators/greenhouse-gas-emissions.html).

The data they provide for download is messy, so we cleaned it up a bit to save time. This often happens.  You can clean the data in R (which is great for reproducibility) but we just did it in Excel and saved the data as a CSV (great for saving time).

Even though the data is cleaned up, it's not "tidy," so we'll tidy it up so that each variable is a column, each observation is a row, and each value has its own cell.
(See [this full description of tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) for more.)  Basically, we'll take the data from being human-readable spreadsheet-style to being more machine-usable.

## Get the data

``` {r}
## Download the data
ghg_raw <- read_csv("https://raw.githubusercontent.com/yorkulibraries/rworkshop/master/ghg-cleaned.csv")

## View the data
## RStudio: click on variable name in upper right
ghg_raw


## We’re using the pipe and the tidyr package here,
## but consider it magic for now.
ghg <- ghg_raw %>% pivot_longer(-year, names_to = "sector", values_to = "emissions")

## View it now
## RStudio: click on variable name in uppper right
ghg
```

## Scatter plots and box plots

It's always nice to start with a scatter plot.  Here we'll do it by sector.

``` {r}
## Always a nice way to start: scatter plot
## Here we’ll do it by sector
ggplot(data = ghg, aes(x = sector, y = emissions)) + geom_point()

## Try just this: no points!
ggplot(data = ghg, aes(x = sector, y = emissions))

## Jitter it up
ggplot(data = ghg, aes(x = sector, y = emissions)) + geom_jitter()

## Tighten the spread
ggplot(data = ghg, aes(x = sector, y = emissions)) + geom_jitter(width = 0.3)

## Box plot (shows first and third quartiles, and median)
## https://ggplot2.tidyverse.org/reference/geom_boxplot.html
ggplot(data = ghg, aes(x = sector, y = emissions)) + geom_boxplot()

## Compare the oil_gas column to the summary data here
ghg %>% filter(sector == "oil_gas") %>% summary()
```

## Exercise 1

* Add `geom_jitter()` to add the points to the box plot.
* What happens if you reverse the order and put `geom_jitter()` before `geom_boxplot()`?
* What happened is because order matters in ggplot commands.
* After trying this, try using geom_boxplot(alpha = 0.5) instead.  The alpha setting controls transparency.

Box plots are useful summaries, but hide the shape of the distribution. For example, if the distribution is bimodal, we would not see it in a box plot. An alternative to the box plot is the violin plot, where the shape (of the density of points) is drawn.

Try `geom_violin()` as well.

Here’s one with some extra new stuff to try if you have time:

``` {r}
ggplot(data = ghg, aes(x = sector, y = emissions)) + 
  geom_jitter(width = 0.3, aes(size = year), alpha = 0.5)
```

## Points and lines

``` {r}
## Now we’ll put year on the x-axis
## And we’ll leave out “data =” because it’s unnecessary
```

## Scatter plot

``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_point()
```

## Just lines

``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_line()
```

That didn’t work!  We need to connect it by sector, and specifying colours is enough for R to know what to do

``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_line(aes(colour = sector))
```

## Lines and points
``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_line(aes(colour = sector)) + geom_point()
```

## Lines and shaped points
``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + 
  geom_line(aes(colour = sector)) + 
  geom_point(aes(shape = sector))
```

## Lines and softer shaped points
``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + 
  geom_line(aes(colour = sector)) + 
  geom_point(aes(shape = sector), alpha = 0.2)
```

## What does this do?

Can you guess from the code before you run it to see the chart?

``` {r}
ghg %>% 
  group_by(sector) %>% 
  mutate(percent_change_to_base = emissions / emissions[1]) %>% 
  ggplot(aes(x = year, y = percent_change_to_base)) + 
  geom_line(aes(colour = sector))
```

## Exercise 2

* Try adding points to this new plot, like before, and also try it with an alpha setting.
* Does this different presentation of the data give you a different understanding of the emissions by sector?

Try picking that last command apart by just looking at the data produced by the `mutate` command.  You can even pipe this to `tail()` to see the bottom.

``` {r}
ghg %>% group_by(sector) %>% mutate(percent_change_to_base = emissions / emissions[1])
```
``` {r}
ghg %>% group_by(sector) %>% mutate(percent_change_to_base = emissions / emissions[1]) %>% tail()
```

## Bar plots

``` {r}
ghg %>% group_by(sector) %>% mutate(percent_change_to_base = emissions / emissions[1]) %>% tail()## Emissions are added up automatically
ggplot(ghg, aes(x = year, y = emissions)) + geom_col()
```

``` {r}
## Fill (with colour) by sector
ggplot(ghg, aes(x = year, y = emissions)) + geom_col(aes(fill = sector))
```

``` {r}
## Break it up by sector, dodge the bars
ggplot(ghg, aes(x = year, y = emissions)) + geom_col(aes(fill = sector), position = "dodge")
```

``` {r}
## Smaller portion of that
ggplot(ghg %>% filter(year >= 2010), aes(x = year, y = emissions)) + geom_col(aes(fill = sector), position = "dodge")
```

## Exercise 3

Try `position = "fill"` instead of "dodge".  (This is different from the fill aesthetic.)  What does that do?

Try

``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_col() + facet_wrap(~ sector)
```

## Customizing the bar plot

``` {r}
## Back to basic bar plot

ggplot(ghg, aes(x = year, y = emissions)) + geom_col(aes(fill = sector))
```

``` {r}
## But the sectors are in alphabetical order
## (which is a very sensible default)
## We want it so biggest is at the bottom

## Arrange the ordering

## First, see what’s going on
ghg %>% filter(year == 2017)
ghg %>% filter(year == 2017) %>% arrange(emissions)
ghg %>% filter(year == 2017) %>% arrange(emissions) %>% pull(sector)

## Now save it and set it
sector_order <- ghg %>% filter(year == 2017) %>% arrange(emissions) %>% pull(sector)
sector_order

## Make a copy; don’t change the original
ghg_ordered <- ghg
ghg_ordered$sector <- factor(ghg_ordered$sector, levels = sector_order)
```

``` {r Replot with reordered sectors}
## Plot with reordered sectors
## Add in labels while we’re at it
ggplot(ghg_ordered, aes(x = year, y = emissions)) + geom_col(aes(fill = sector))
```

``` {r Add titles and labels}
## Add titles and labels
ggplot(ghg_ordered, aes(x = year, y = emissions)) + 
  geom_col(aes(fill = sector)) + 
  labs (title = "Greenhouse gas emissions by economic sector, Canada, 1990 to 2017", 
        subtitle = "Megatonnes of carbon dioxide equivalent", 
        x = "", 
        y = "", 
        caption = "Data from Environment and Climate Change Canada, 2019", 
        fill = "")
```

``` {r}
## Change colours and the overall look
ggplot(ghg_ordered, aes(x = year, y = emissions)) + 
  geom_col(aes(fill = sector)) + 
  labs (title = "Greenhouse gas emissions by economic sector, Canada, 1990 to 2017", 
        subtitle = "Megatonnes of carbon dioxide equivalent", 
        x = "", 
        y = "", 
        caption = "Data from Environment and Climate Change Canada, 2019", 
        fill = "") + 
  scale_fill_brewer(palette = "Paired") + 
  theme_minimal()
```

``` {r}
## Fix the names of the sectors
levels(ghg_ordered$sector) <- c("Waste and others", 
                                "Agriculture", 
                                "Heavy industry", 
                                "Electricity", 
                                "Buildings", 
                                "Transportation", 
                                "Oil and gas")
```

``` {r}
## Now rerun the previous plot
ggplot(ghg_ordered, aes(x = year, y = emissions)) + 
  geom_col(aes(fill = sector)) + 
  labs (title = "Greenhouse gas emissions by economic sector, Canada, 1990 to 2017", 
        subtitle = "Megatonnes of carbon dioxide equivalent", 
        x = "", 
        y = "", 
        caption = "Data from Environment and Climate Change Canada, 2019", 
        fill = "") + 
  scale_fill_brewer(palette = "Paired") + 
  theme_minimal()
```


## Exercise 4

Try customizing the plot further by setting the x- and y-axis labels and editing the titles and the caption.

Try adding `position = "fill"` to `geom_col(aes(fill = sector))` (after the aesthetic, before the final brackets).

Try it with `geom_line(aes(colour = sector))` instead of the `geom_col()`.  How does this make the data look different?
