---
title: Introduction to Analyzing and Visualizing Data with R
output: html_document
---

# About R Markdown and RStudio

*To be done*.  Show some exciting magic.

https://rmarkdown.rstudio.com/

RStudio and R Markdown are a great combination for working, keeping notes, reproducible research and collaborating.  

(You may also want to look at Jupyter Notebooks (https://jupyter.org/), which can be used with R, Julia or Python.)

Review the RStudio panes.  Explain about how some output will appear in this buffer, but there's also output in the console below.  

# Setup

Please run these two commands.  It will make things easier right away, but we'll explain why a little later.

This installs the *tidyverse* package.  (You only need to do this once on your machine.  This is like installing a program.)

``` {r Setup}
## Anything starting with ## is a comment, not code
## Remove them if you want to run the line

install.packages("tidyverse") ## If not already done
```

And this loads it in so we can use it.  (You'll need to do this every time you start up RStudio fresh.  This is like using an installed program.)

``` {r Setup}
library(tidyverse) ## Will load ggplot2 and others
```

# Get the data 

We'll be using two different data sets so we can show differences in how qualitative and quantitative data are handled, what to do about missing values, and so on.

## Survey data 

Use # to comment in scripts and R markdown.

``` {r Load data}
# Load the data for this lesson by using the read_csv function (from the tidyverse) and assigning the dataset to an object named "surveys"
surveys <- read_csv("http://kbroman.org/datacarp/portal_data_joined.csv")
```

## Greenhouse gas emissions

Another example data set we'll use is [greenhouse gas emissions data from Environment and Climate Change Canada](https://www.canada.ca/en/environment-climate-change/services/environmental-indicators/greenhouse-gas-emissions.html).

The data they provide for download is messy, so we cleaned it up a bit to save time. 

BILL: Change this so they load in tidy data.

``` {r}
## Download the data
ghg_raw <- read_csv("https://raw.githubusercontent.com/yorkulibraries/rworkshop/master/ghg-cleaned.csv")

## View the data
## RStudio: click on variable name in upper right
ghg_raw


## We’re using the pipe and the tidyr package here,
## but consider it magic for now.
ghg <- ghg_raw %>% pivot_longer(-year, names_to = "sector", values_to = "emissions")

## View it now
## RStudio: click on variable name in uppper right
ghg
```

# Analyzing data

## Using R like a calculator

``` {r}
# Using R like a calculator
2020 - 1971  # How old is Justin Trudeau this year?

9.95 * 1.13  # Cost of a paperback after tax

pi * 2^2  # Area of a circle of radius 2

```

## Inspecting data frames

``` {r}
# head: shows the top few rows
head(surveys)
```

``` {r}
# tail: shows the bottom few rows
tail(surveys)
```

``` {r}
# str shows the structure
str(surveys)
```

##  Challenge 1

(Explain how to make a new little code block as a workspace.)

Study the output of str(surveys). How are missing values being treated?

Other useful ways of inspecting data frames:

dim(surveys)
ncol(surveys)
nrow(surveys)
names(surveys)
colnames(surveys)
rownames(surveys)
summary(surveys)

## Indexing

Data frames have rows and columns, and there are various ways to get at one or the other or a specific cell.  When indexing, rows go first and columns second: you count down and then over.

``` {r}
surveys[1, 1]
surveys[1, 7]
surveys[10001, 7]
surveys[2, ]
surveys[2,]  # don't _need_ the spaces
surveys[,7]  # Helpful things with tibbles and RStudio make it easier to inspect this
```

## Vectors

``` {r}
sex <- surveys[,7]
sex <- surveys[, "sex"]
sex <- surveys$sex
sex <- surveys[["sex"]]
```

``` {r}
# Pull out a single value from a vector
sex[1]
sex[10001]
```

``` {r}
# Create a vector 
c(1, 4, 6)
```

``` {r}
# Pull out multiple values
sex[c(1,4,6)] # Why are 4 and 6 not seen?
```

``` {r}
sex[c(1, 10001)] # <- pull out 1st and 10001th.
```

``` {r}
# More ways to create vectors
1:10
10:1
sex[1:10]
```

#every 2nd value
seq(1, 10, by=2)
sex[seq(1,10, by=2)]

#seq function
?seq # help
seq(1, 11)
seq(1, 11, 2)
seq(1, 11, by=2)
seq(to=11, from=1, by=2)

#first ten rows of surveys
surveys[1:3,]
surveys[10001:10003,]
surveys[5:7, 1:7]

## Challenge 2

The function nrow() on a data.frame returns the number of rows.

Use nrow(), in conjuction with seq() to create a new data frame called surveys_by_10 that includes every 10th row of the survey data frame starting at row 10 (10, 20, 30, .)

``` {r}
# How would you  make sequence that goes up by tens?
# How would you use that to extract just those rows from surveys?
# Try some code here.  Edit and rerun as you work on it.

```

## Dealing with missing data

``` {r}
heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
na.omit (heights)
```

surveys_noblanks <- read.csv("http://kbroman.org/datacarp/portal_data_joined.csv", na.strings="")

## What is the Tidyverse?

Indeed, what?  Brief explanation.  Get ready for pipes.

## Filtering

``` {r}
# select some columns
selected_col <- select(surveys, sex, species_id, plot_type, weight)
head(selected_col)

#filter out some rows
selected_row <- filter(surveys, year == 2002)
head(selected_row)

#filter out some rows
selected_row <- filter(surveys, year == 2002, weight>78)
head(selected_row)

selected_row <- filter(surveys, sex == "F", weight>78)
head(selected_row)

#pipe operator
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

selected_rows <- filter(surveys, weight<5)
result <- select(selected_rows, species_id, sex, weight)
```

## Challenge 3

Using pipes, subset the data to include individuals collected before 1995, and retain the columns year, sex, and weight.

# Visualizing data

## Introduction

The goal here is to show how the [ggplot2](https://ggplot2.tidyverse.org/) system works (combining commands to build up a chart piece by piece), to show a variety of examples, and to recreate an existing chart in R.

## Data: Canadian greenhouse gas emissions

We already have the data.

TODO:  Go back and explain about wide versus tall data and how we tidied it up.

``` {r}
## Download the data
ghg_raw <- read_csv("https://raw.githubusercontent.com/yorkulibraries/rworkshop/master/ghg-cleaned.csv")

## View the data
## RStudio: click on variable name in upper right
ghg_raw


## We’re using the pipe and the tidyr package here,
## but consider it magic for now.
ghg <- ghg_raw %>% pivot_longer(-year, names_to = "sector", values_to = "emissions")

## View it now
## RStudio: click on variable name in uppper right
ghg
```

## Scatter plots and box plots

It's always nice to start with a scatter plot.  Here we'll do it by sector.

TODO:  Break these up into one bit of code per chunk.

``` {r}
## Always a nice way to start: scatter plot
## Here we’ll do it by sector
ggplot(data = ghg, aes(x = sector, y = emissions)) + geom_point()

## Try just this: no points!
ggplot(data = ghg, aes(x = sector, y = emissions))

## Jitter it up
ggplot(data = ghg, aes(x = sector, y = emissions)) + geom_jitter()

## Tighten the spread
ggplot(data = ghg, aes(x = sector, y = emissions)) + geom_jitter(width = 0.3)

## Box plot (shows first and third quartiles, and median)
## https://ggplot2.tidyverse.org/reference/geom_boxplot.html
ggplot(data = ghg, aes(x = sector, y = emissions)) + geom_boxplot()

## Compare the oil_gas column to the summary data here
ghg %>% filter(sector == "oil_gas") %>% summary()
```

## Challenge

* Add `geom_jitter()` to add the points to the box plot.
* What happens if you reverse the order and put `geom_jitter()` before `geom_boxplot()`?
* What happened is because order matters in ggplot commands.
* After trying this, try using geom_boxplot(alpha = 0.5) instead.  The alpha setting controls transparency.

Box plots are useful summaries, but hide the shape of the distribution. For example, if the distribution is bimodal, we would not see it in a box plot. An alternative to the box plot is the violin plot, where the shape (of the density of points) is drawn.

Try `geom_violin()` as well.

Here’s one with some extra new stuff to try if you have time:

``` {r}
ggplot(data = ghg, aes(x = sector, y = emissions)) + 
  geom_jitter(width = 0.3, aes(size = year), alpha = 0.5)
```

## Points and lines

``` {r}
## Now we’ll put year on the x-axis
## And we’ll leave out “data =” because it’s unnecessary
```

## Scatter plot

``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_point()
```

## Just lines

``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_line()
```

That didn’t work!  We need to connect it by sector, and specifying colours is enough for R to know what to do

``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_line(aes(colour = sector))
```

## Lines and points
``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_line(aes(colour = sector)) + geom_point()
```

## Lines and shaped points
``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + 
  geom_line(aes(colour = sector)) + 
  geom_point(aes(shape = sector))
```

## Lines and softer shaped points
``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + 
  geom_line(aes(colour = sector)) + 
  geom_point(aes(shape = sector), alpha = 0.2)
```

## What does this do?

Can you guess from the code before you run it to see the chart?

``` {r}
ghg %>% 
  group_by(sector) %>% 
  mutate(percent_change_to_base = emissions / emissions[1]) %>% 
  ggplot(aes(x = year, y = percent_change_to_base)) + 
  geom_line(aes(colour = sector))
```

## Challenge

* Try adding points to this new plot, like before, and also try it with an alpha setting.
* Does this different presentation of the data give you a different understanding of the emissions by sector?

Try picking that last command apart by just looking at the data produced by the `mutate` command.  You can even pipe this to `tail()` to see the bottom.

``` {r}
ghg %>% group_by(sector) %>% mutate(percent_change_to_base = emissions / emissions[1])
```
``` {r}
ghg %>% group_by(sector) %>% mutate(percent_change_to_base = emissions / emissions[1]) %>% tail()
```

## Bar plots

``` {r}
ghg %>% group_by(sector) %>% mutate(percent_change_to_base = emissions / emissions[1]) %>% tail()## Emissions are added up automatically
ggplot(ghg, aes(x = year, y = emissions)) + geom_col()
```

``` {r}
## Fill (with colour) by sector
ggplot(ghg, aes(x = year, y = emissions)) + geom_col(aes(fill = sector))
```

``` {r}
## Break it up by sector, dodge the bars
ggplot(ghg, aes(x = year, y = emissions)) + geom_col(aes(fill = sector), position = "dodge")
```

``` {r}
## Smaller portion of that
ggplot(ghg %>% filter(year >= 2010), aes(x = year, y = emissions)) + geom_col(aes(fill = sector), position = "dodge")
```

## Challenge

Try `position = "fill"` instead of "dodge".  (This is different from the fill aesthetic.)  What does that do?

Try

``` {r}
ggplot(ghg, aes(x = year, y = emissions)) + geom_col() + facet_wrap(~ sector)
```

## Customizing the bar plot

``` {r}
## Back to basic bar plot

ggplot(ghg, aes(x = year, y = emissions)) + geom_col(aes(fill = sector))
```

``` {r}
## But the sectors are in alphabetical order
## (which is a very sensible default)
## We want it so biggest is at the bottom

## Arrange the ordering

## First, see what’s going on
ghg %>% filter(year == 2017)
ghg %>% filter(year == 2017) %>% arrange(emissions)
ghg %>% filter(year == 2017) %>% arrange(emissions) %>% pull(sector)

## Now save it and set it
sector_order <- ghg %>% filter(year == 2017) %>% arrange(emissions) %>% pull(sector)
sector_order

## Make a copy; don’t change the original
ghg_ordered <- ghg
ghg_ordered$sector <- factor(ghg_ordered$sector, levels = sector_order)
```

``` {r Replot with reordered sectors}
## Plot with reordered sectors
## Add in labels while we’re at it
ggplot(ghg_ordered, aes(x = year, y = emissions)) + geom_col(aes(fill = sector))
```

``` {r Add titles and labels}
## Add titles and labels
ggplot(ghg_ordered, aes(x = year, y = emissions)) + 
  geom_col(aes(fill = sector)) + 
  labs (title = "Greenhouse gas emissions by economic sector, Canada, 1990 to 2017", 
        subtitle = "Megatonnes of carbon dioxide equivalent", 
        x = "", 
        y = "", 
        caption = "Data from Environment and Climate Change Canada, 2019", 
        fill = "")
```

``` {r}
## Change colours and the overall look
ggplot(ghg_ordered, aes(x = year, y = emissions)) + 
  geom_col(aes(fill = sector)) + 
  labs (title = "Greenhouse gas emissions by economic sector, Canada, 1990 to 2017", 
        subtitle = "Megatonnes of carbon dioxide equivalent", 
        x = "", 
        y = "", 
        caption = "Data from Environment and Climate Change Canada, 2019", 
        fill = "") + 
  scale_fill_brewer(palette = "Paired") + 
  theme_minimal()
```

``` {r}
## Fix the names of the sectors
levels(ghg_ordered$sector) <- c("Waste and others", 
                                "Agriculture", 
                                "Heavy industry", 
                                "Electricity", 
                                "Buildings", 
                                "Transportation", 
                                "Oil and gas")
```

``` {r}
## Now rerun the previous plot
ggplot(ghg_ordered, aes(x = year, y = emissions)) + 
  geom_col(aes(fill = sector)) + 
  labs (title = "Greenhouse gas emissions by economic sector, Canada, 1990 to 2017", 
        subtitle = "Megatonnes of carbon dioxide equivalent", 
        x = "", 
        y = "", 
        caption = "Data from Environment and Climate Change Canada, 2019", 
        fill = "") + 
  scale_fill_brewer(palette = "Paired") + 
  theme_minimal()
```


## Challenge

Try customizing the plot further by setting the x- and y-axis labels and editing the titles and the caption.

Try adding `position = "fill"` to `geom_col(aes(fill = sector))` (after the aesthetic, before the final brackets).

Try it with `geom_line(aes(colour = sector))` instead of the `geom_col()`.  How does this make the data look different?

# Exporting from RStudio

Knitting; exporting to HTML and PDF.  

A bit about using RStudio as a living notebook for research and notes, etc.




